---
title: "Making choices"
author: "Luna L Sanchez Reyes"
date: "2022-11-10"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Logical and conditional statements

Both of these return `TRUE` or `FALSE` values, which are values of "logical" types or class.

Some examples of logical statements (usually called comparison between values):
```{r}
1 == 1
1 > 1
1 < 1
1 >= 1
1 <= 1

"a" == "a"
"a" < "a"
"a" < "b"
"b" < "c"
"apple" < "banana"
"apple" < "airplane"

c(1, 2, 3, 4, 1, 1) == 1

letters
letters < "m"
letters[letters < "m"]
LETTERS

TRUE == FALSE
```

Load data from the UHURU data set:

```{r}
surveys <- read.csv("../data-raw/surveys.csv")
species <- read.csv("../data-raw/species.csv")

colnames(surveys)
```
Combine the two tables:
```{r}
library(dplyr)
intersect(colnames(surveys), colnames(species))
combined <- dplyr::inner_join(surveys, species, by = "species_id")

colnames(combined)
```
Extract the "year" column:
```{r}
str(combined)
dim(combined)
combined$year
combined[, "year"]
combined["year"]
combined[4]

combined$year == 1977
combined$genus == "Dipodomys"
data.frame(combined$genus, combined$genus == "Dipodomys")
```

A vector that tests speciess equal to "merriami":

```{r}
combined$species == "merriami"
data.frame(combined$species, combined$species == "merriami")
```

Some examples of conditional statements (logical statement AND logical statement we use `&`, OR `|`):

```{r}
1 == 1 & 1 < 2
1 == 1 & 1 == 2
1 == 1 | 1 == 2
```
```{r}
combined$year == 1977 & combined$genus == "Dipodomys" & combined$species == "merriami"
```

## The basic `if` statement

Complete the if statement so that if age_class is equal to “sapling” it sets y <- 10.
What is the value of y?

```{r}
age_class <-  "sapling"
 if (age_class == "sapling"){
  y <- 10
 }
 y
```

`y` has a value of 10.

## The if/else statement

Complete the if statement so that if age_class is equal to “sapling” it sets y <- 10 and if age_class is equal to “seedling” it sets y <- 5.
```{r}
age_class <-  "seedling"
 
if (age_class == "sapling"){
    y <- 10 
  } else {
    y <- 5
}

y
```

## The if, else if, else statement

This statement handles three or more choices:

Complete the if statement so that if age_class is equal to “sapling” it sets y <- 10 and if age_class is equal to “seedling” it sets y <- 5 and if age_class is something else then it sets the value of y <- 0.

```{r}
age_class <-  "sapling"
if (age_class == "sapling"){
  y <- 10 
} else if (age_class == "seedling") {
  y <- 5
} else if (age_class == "sapling") {
  y <- "test"
} else if (age_class != "seedling") {
  y <-0
}

y
```
The most straightforward, less confusing way of doing the statement:

```{r}

if (age_class == "sapling"){
  y <- 10 
} else if (age_class == "seedling") {
  y <- 5
} else {
  y <-0
}
```

## Using if else inside functions

Convert your conditional statement from the last exercise into a function called get_y, that takes age_class as an argument and returns y.

```{r}
name_for_your_function <- function(argument1, argument2, argument3) {
  Body of the function
}
```

```{r}
get_y <- function(age_class = "adult") {
  if (age_class == "sapling"){
    y <- 10 
  } else if (age_class == "seedling") {
    y <- 5
  } else {
    y <-0
  }
  return(y)
}

```

Question 2: run the function using the values "sapling”,
“seedling”,
“adult”,
“mature”,
“established”.

```{r}
get_y("sapling")
get_y(age_class = "seedling")
get_y("adult")
get_y("mature")
get_y("established")
```

Exercise 6:

Question 1: Create a new function `get_mass_from_length_by_name()` that:
takes two arguments, the `length` and the `name` of the dinosaur group.
Inside this function use `if`/`else if`/`else` statements to check if the name is one of the following values and if so set a and b to the appropriate values (from Seebacher 2001):

- Stegosauria: a = 10.95 and b = 2.64.
- Theropoda: a = 0.73 and b = 3.63.
- Sauropoda: a = 214.44 and b = 1.46.
- If the name is not any of these values, the function should return NA. 

Once the function has chosen the correct a and b values, have it run `get_mass_from_length()` with the appropriate values and return the estimated mass.


Define the function `get_mass_from_length()` from the previous class:
```{r}
get_mass_from_length <- function(length, a, b){
  mass <- a * length ^ b
  return(mass)
}
```

Create the function `get_mass_from_length_by_name()`:
```{r}
get_mass_from_length_by_name <- function(length, dinosaur_name) {
  if (dinosaur_name == "Stegosauria") {
    a <- 10.95
    b <- 2.64
  } else if (dinosaur_name == "Theropoda") {
    a <- 0.73
    b <- 3.63
  } else if (dinosaur_name == "Sauropoda") {
    a <- 214.44
    b <- 1.46
  } else {
    message("No known estimation for ", dinosaur_name)
    return(NA)
  }
  mass <- get_mass_from_length(length = length, a = a, b = b)
  return(mass)
}

get_mass_from_length_by_name(dinosaur_name = "Luna", length = 10)
```

```{r}
get_mass_from_length_by_name = function(length,name) {
   if (name == "Stegosauria") {
     a <- 10.95
     b <- 2.64 
   } else if (name == "Theropoda") {
     a <- 0.73
     b <- 3.63
   } else if (name == "Sauropoda") {
     a <- 214.44
     b <- 1.46
   } else { 
    message("No known estimation for", name)
   return(NA)
   }
 mass <- get_mass_from_length(length = length, a = a, b = b )
 return(mass)
}
```



