---
title: "Creating functions"
author: "Luna L Sanchez Reyes"
date: "2022-11-08"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to creating functions

Based on your experience with coding so far and after watching the videos, list three reason why creating our own functions is good practice for data science.

- It allows reusing code for other parts of a project or a future project
- It is less error prone that copy pasting code
- Code is more organized for you and others, so it is easier to read
- Allows deeper understanding of what your code is doing
- Useful to add data to data frames
- Allows getting specific values and requirements
- Invites you to be intentional on the code that you are writing
_ Allows to be more in control of the outputs

## Exercise 1

Part 1:
```{r}
convert_pounds_to_grams <- function(pounds) {
    grams = 453.6 * pounds
    return(grams)
}

convert_pounds_to_grams(10)
```

Part 2:

```{r}
convert_pounds_to_grams <- function(pounds = 3.75) {
    grams = 453.6 * pounds
    return(grams)
}

convert_pounds_to_grams() # this requires a default value in the function
# without default value
convert_pounds_to_grams(3.75)
convert_pounds_to_grams(pounds = 3.75)
```

## Exercise 2

Part 1

```{r}
get_mass_from_length_theropoda <- function(length){
  mass <- 0.73 * length ^ 3.63
  return(mass)
}
get_mass_from_length_theropoda(16)
```

Part 2
Create a new version of this function called get_mass_from_length() that takes length, a and b as arguments and uses the following code to estimate the mass mass <- a * length ^ b. Use this function to estimate the mass of a Sauropoda (a = 214.44, b = 1.46) that is 26 m long.
```{r, eval = FALSE}
get_mass_from_length <- function(length, a, b){
  mass <- a * length ^ b
  return(mass)
}

a = 214.44
a
a <- 214.44
a
b <- 1.46
get_mass_from_length(length = 26, a, b)
```

*** Investigate why R is taking variables from global environment even inside the function enviornment ***


## Exercise 3

Rewrite your get_mass_from length() function from the precious exercise so that its arguments have default values of a = 39.9 and b = 2.6 (the average values from Seebacher 2001).

1. Use this function to estimate the mass of a Sauropoda (a = 214.44, b = 1.46) that is 22 m long (by setting a and b when calling the function).
2. Use this function to estimate the mass of a dinosaur from an unknown taxonomic group that is 16m long. Only pass the function length, not a and b, so that the default values are used.

```{r}
get_mass_from_length <- function(length, a = 39.9, b = 2.6){
  mass <- a * length ^ b
  return(mass)
}

get_mass_from_length(length = 22, a = 214.44, b = 1.46)
get_mass_from_length(length = 16)
```


## Exercise 4

1. Write a function called `convert_kg_to_pounds()` that converts kilograms into pounds using the formula `pounds = 2.205 * kg`.

```{r}
convert_kg_to_pounds <- function(kg) {
  pounds <- 2.205 * kg
  return(pounds)
}

convert_kg_to_pounds(5)
```

2. Combine that function and your get_mass_from_length() function from the previous exercise to estimate the weight, in pounds, of a 12 m long Stegosaurus with a = 10.95 and b = 2.64 (The estimated a and b values for Stegosauria from Seebacher 2001).

Using the pipe:
```{r}
library(magrittr)
get_mass_from_length(length = 12, a = 10.95, b = 2.64) %>% 
  convert_kg_to_pounds()
```