---
title: "Loops"
author: "Luna L Sanchez Reyes"
date: "2022-11-17"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
sum(1, 2, 3, 4, 5)
sum(19, 11, 04, 27)
sum(37, 456, 378)
```
Exercise 1:

Part 1.
```{r}
numbers <- c(1, 2, 3, 4, 5)
numbers <- 1:5
numbers <- seq(1, 5, 0.3)
read.csv(file = "../data-raw/species.csv")
# mass <- a * length ^ b
for (k in 1:5) {
  print(x = k)
  print("k")
}
```

```{r}
our_names <- c("Azul", "Jai", "Marcos", "Luna")
print(our_names)

for (items in our_names) {
  print(items)
}
```

Part 2. Write a for loop that loops over the following vector and prints out the mass in kilograms (mass_kg = 2.2 * mass_lb)

```{r}
mass_lbs <- c(2.2, 3.5, 9.6, 1.2)

for (items in mass_lbs) {
  print(items)
  mass_kg <- 2.2 * items
  print(mass_kg)
}


```

Part 3. Complete the code below so that it prints out the name of each bird one line at a time.

```{r}
birds = c('robin', 'woodpecker', 'blue jay', 'sparrow')

for (i in 1:length(birds)){
  print(birds[i])
}
```

Part 4. Complete the code below so that it stores one area for each radius.

```{r}
radius <- c(1.3, 2.1, 3.5)

areas <- vector(mode = "numeric", length = length(radius))

for (i in 1:length(radius)){
  print(areas)
  areas[i] <- pi * radius[i] ^ 2
  print(areas)
}
areas

```

Part 5. Complete the code below to calculate an area for each pair of lengths and widths, store the areas in a vector, and after they are all calculated print them out:

```{r}
lengths = c(1.1, 2.2, 1.6)
widths = c(3.5, 2.4, 2.8)
areas <- vector(length = length(lengths), mode = "numeric")
length(areas)
areas
for (i in 1:length(lengths)) {
  areas[i] <- lengths[i] * widths[i]
}
areas

```

Exercise 2.

Part 1. Write a function `mass_from_length()` that uses the equation `mass <- a * length^b` to estimate the size of a dinosaur from its length. This function should take two arguments, `length` and `species.` For each of the following inputs for species, use the given values of a and b for the calculation from Seebacher 2001:

For Stegosauria: a = 10.95 and b = 2.64
For Theropoda: a = 0.73 and b = 3.63
For Sauropoda: a = 214.44 and b = 1.46
For any other value of species: a = 25.37 and b = 2.49

```{r}
mass_from_length <- function(length, species) {
  if (species == "Stegosauria") {
    a <- 10.95
    b <- 2.64
  } else if (species == "Theropoda") {
    a <- 0.73
    b <- 3.63
  } else if (species == "Sauropoda") {
    a <- 214.44
    b <- 1.46
  } else {
    a <- 25.37
    b <- 2.49
  }
  mass <- a * length ^ b
  return(mass)
}
```

Part 2. Use this function and a for loop to calculate the estimated mass for each dinosaur, store the masses in a vector, and after all of the calculations are complete show the first few items in the vector using head().

```{r}
dino <- read.csv(file = "../data-raw/dinosaur_lengths.csv")
head(dino)

library(dplyr)
dino %>% 
  rowwise() %>% 
  mutate(masses = mass_from_length(lengths, species))

masses <- vector(mode = "numeric", length = nrow(dino))
for (i in seq(nrow(dino))) {
  
  masses <- mass_from_length()
}
```


Part 3. Add the results in the vector back to the original data frame. Show the first few rows of the data frame using `head()`.

Part 4. Calculate the mean mass for each species using `dplyr.`










