---
title: "Node labels on trees"
author: "Luna L Sanchez Reyes"
date: "2022-11-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The data set

Data that comes with an R package.
Working with the `TDbook` package.

Load the packages and the data package `TDbook`:

```{r}
install.packages("TDbook")
library(TDbook)
library(ggtree)
library(tidytree)
```

Load the data from the data package, using the `data()` function.
The data that we are loading is 
- tree_boots
- df_tip_data
- df_inode_data

```{r}
data("tree_boots", "df_inode_data", "df_tip_data")
tree_boots
```

Exercise
1. What is the class of this three objects
2. How many elements does `tree_boots` has
3. How many rows does `df_tip_data` and `df_inode_data` have.
4. Compare the row numbers of both objects to the lengths of the `$tip.label` and `$node.label` elements of `tree_boots`.
5. Join the tree object with the data table that contains tip data and create an object called `tree_joined`.

```{r}
class(tree_boots)
class(df_tip_data)
class(df_inode_data)
typeof(tree_boots)
typeof(df_tip_data)
typeof(df_inode_data)
nrow(df_tip_data) # 7
nrow(df_inode_data) #6
length(df_tip_data) #10 when we are working with a data frame, length gives us the number of columns
length(df_tip_data$Newick_label)
length(df_tip_data$trophic_habit)
length(df_inode_data$newick_label)
length(df_inode_data) #6
colnames(df_inode_data)
colnames(df_tip_data)
length(tree_boots)
names(tree_boots)
```

Getting the Length of tip.label and node.label in tree_boots:

```{r}
nrow(df_tip_data) #7
tree_boots$tip.label %>% length() # 7
nrow(df_inode_data) # 6
tree_boots$node.label %>% length() # 6
```

Joining tree_boots and df_tip_data:

```{r}
librray(tidytree) # not the one from dplyr package
colnames(df_tip_data)
df_tip_data$Newick_label
tree_boots$tip.label
# rename the "Newick_label" column as "label":
df_mutated <- mutate(df_tip_data, label = Newick_label)
colnames(df_mutated)
df_mutated$Newick_label
df_mutated$label
left_join(tree_boots, df_mutated, by = "label")

df_tip_data

colnames(df_tip_data) -> my_colnames
my_colnames
class(my_colnames)
typeof(my_colnames)
is.vector(my_colnames)
my_colnames[1]
my_colnames[1] <- "column 1"
my_colnames[10] <- "taxonomy"

colnames(df_tip_data)[10] <- "taxonomy"
colnames(df_tip_data)[1] <- "label"
df_tip_data

# try to join again tree_boots and df_tip_data:

tree_joined <- left_join(tree_boots, df_tip_data)
tree_joined
tree_boots
```

```{r}
tree_joined@phylo
tree_joined@extraInfo$mass_in_kg
```


### Plot node labels intro a tree

```{r}
df_tip_data
df_inode_data
```

Join tree_boots to df_inode_data. Hint: Pay attention to the names of the columns, first change the name of the "newick_label" column to "label":

```{r}
colnames(df_inode_data)[1] <- "label"
tree_data <- left_join(tree_boots, df_inode_data)
tree_data@extraInfo
```

```{r}
ggtree(tree_data) + 
  geom_tiplab(fontface = "italic") + 
  xlim(0, 5) +
  geom_nodelab(mapping = aes(label = vernacularName, color = vernacularName))
```

Exercises:
Add node labels to your two tree plots of Portal tree data with average weight and hindfoot length. Use the column "taxa" both as label and fill color.
1. Join portal_tree and the dataframe with average weights and hindfoot lengths, and taxa data.
2. Plot it
3. Add tip labels 
4. Add node labels 
5. Colored by taxa

```{r}
species <- read.csv("../data-raw/species.csv")
surveys <- read.csv("../data-raw/surveys.csv")
library(dplyr)
taxonomy_means <- full_join(surveys, species) %>% 
  filter(!is.na(weight) & !is.na(hindfoot_length)) %>% 
  group_by(species_id) %>% 
  summarise(mean_weight = mean(weight), mean_hindfoot = mean(hindfoot_length))

portal_taxonomy <- read.csv("../data-raw/portal-species-taxonomy.csv")

portal_taxonomy_means <- full_join(portal_taxonomy, taxonomy_means)
library(tidytree)
portal_tree <- read.tree(file = "../data-raw/portal-tree.tre")
portal_joined <- tidytree:::left_join.phylo(portal_tree, portal_taxonomy_means)
portal_joined
```

```{r}
ggtree(portal_joined) + 
  geom_nodelab()
```