---
title: "portal phylogenies"
author: "Luna L Sanchez Reyes"
date: "2022-10-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Add an intro to the newick format

# The package `ape`

```{r}
# install.packages("ape")
library(ape)
```

To read a newick tree from a local file:

```{r}
portal_tree <- read.tree(file = "../data-raw/portal-tree.tre")
portal_tree
```

Read a newick from a URL

```{r}
small_tree <- read.tree(file = "http://ape-package.ird.fr/APER/APER2/primfive.tre")
small_tree
```

# The structure of a tree in R

```{r}
class(portal_tree)
surveys <- read.csv(file = "../data-raw/surveys.csv")
class(surveys)
length(portal_tree)
length(surveys)
colnames(surveys)
colnames(portal_tree)
names(portal_tree)
portal_tree$Nnode
portal_tree["tip.label"]
head(portal_tree[[3]])
summary(portal_tree)
str(portal_tree)
class(portal_tree$edge)
head(portal_tree$edge)
```

```{r}
str(surveys)
```

## Plotting a phylogenetic tree

```{r}
plot.phylo(x = portal_tree)
```

An extension of ggplot for phylogenetic visualization: `ggtree`
To install packages from th Bioconductor repo, we need a CRAN package called BiocManager

```{r, eval=FALSE}
install.packages("BiocManager")
install("ggtree")
```

```{r}
library(BiocManager)
library(ggtree)
ggtree(portal_tree)
```

This is equivalent to:

```{r}
ggplot(portal_tree, aes(x, y)) +
  geom_tree() +
  theme_tree()
```

Add a scale use the function geom_treescale():

```{r}
ggtree(portal_tree) +
  geom_treescale()
```
Exercise: Plot the small tree of five species of primates and include a scale. What is the difference in structure between the two trees?
```{r}
ggtree(small_tree) +
  geom_treescale()
```
```{r}
portal_tree
small_tree
names(portal_tree)
```
```{r}
names(small_tree)
```
- A difference in number of tips (43 vs 5)
- An `edge.length` in `small_tree`, but it has no node labels

```{r}
small_tree$edge.length
small_tree$node.label
head(portal_tree$node.label)
```

## Add tip labels and node labels

Because a plot is nothing with no labels:

```{r}
ggtree(portal_tree) +
  geom_treescale() +
  geom_tiplab(size = 1, color = "purple", fontface = "italic")
```


Add a limit to the plot so we can see the labels fully:

```{r}
ggtree(portal_tree) +
  geom_tiplab(size = 1, color = "purple", fontface = "italic") +
  xlim(NA,200)
```

```{r}
library(ape)
branching.times(small_tree)
branching.times(portal_tree)
```

# Homework: Make a plot for small_tree with scale, tip labels, and an appropriate xlim size to display the species names fully