---
title: "portal phylogenies"
author: "Luna L Sanchez Reyes"
date: "2022-10-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Add an intro to the newick format

# The package `ape`

```{r}
# install.packages("ape")
library(ape)
```

To read a newick tree from a local file:

```{r}
portal_tree <- read.tree(file = "../data-raw/portal-tree.tre")
portal_tree
```

Read a newick from a URL

```{r}
small_tree <- read.tree(file = "http://ape-package.ird.fr/APER/APER2/primfive.tre")
small_tree
```

# The structure of a tree in R

```{r}
class(portal_tree)
surveys <- read.csv(file = "../data-raw/surveys.csv")
class(surveys)
length(portal_tree)
length(surveys)
colnames(surveys)
colnames(portal_tree)
names(portal_tree)
portal_tree$Nnode
portal_tree["tip.label"]
head(portal_tree[[3]])
summary(portal_tree)
str(portal_tree)
class(portal_tree$edge)
head(portal_tree$edge)
```

```{r}
str(surveys)
```

## Plotting a phylogenetic tree

```{r}
plot.phylo(x = portal_tree)
```

An extension of ggplot for phylogenetic visualization: `ggtree`
To install packages from th Bioconductor repo, we need a CRAN package called BiocManager

```{r, eval=FALSE}
install.packages("BiocManager")
install("ggtree")
```

```{r}
library(BiocManager)
library(ggtree)
ggtree(portal_tree)
```

This is equivalent to:

```{r}
ggplot(portal_tree, aes(x, y)) +
  geom_tree() +
  theme_tree()
```

Add a scale use the function geom_treescale():

```{r}
ggtree(portal_tree) +
  geom_treescale()
```
Exercise: Plot the small tree of five species of primates and include a scale. What is the difference in structure between the two trees?
```{r}
ggtree(small_tree) +
  geom_treescale()
```
```{r}
portal_tree
small_tree
names(portal_tree)
```
```{r}
names(small_tree)
```
- A difference in number of tips (43 vs 5)
- An `edge.length` in `small_tree`, but it has no node labels

```{r}
small_tree$edge.length
small_tree$node.label
head(portal_tree$node.label)
```

## Add tip labels and node labels

Because a plot is nothing with no labels:

```{r}
ggtree(portal_tree) +
  geom_treescale() +
  geom_tiplab(size = 1, color = "purple", fontface = "italic")
```


Add a limit to the plot so we can see the labels fully:

```{r}
ggtree(portal_tree) +
  geom_tiplab(size = 1, color = "purple", fontface = "italic") +
  xlim(NA,200)
```

```{r}
library(ape)
branching.times(small_tree)
branching.times(portal_tree)

```

# Homework: Make a plot for small_tree with scale, tip labels, and an appropriate xlim size to display the species names fully

```{r}
ggtree(small_tree) +
  geom_tiplab(size = 5, color = "gray", fontface = "italic") +
  xlim(NA,1.2) +
  geom_nodelab(size = 3, color = "blue")
```

```{r}
library(ggtree)
ggtree(portal_tree, layout = "fan", open.angle = 45) -> plot1
```
```{r}
# install.packages("aplot")

library(aplot)
plot_list(plot1, plot1, labels = c("Circular Tree of Portal Species", "The same"), tag_size = 7)
```
Exercise: Tree representation

Part 1:
```{r}
p1 <- ggtree(portal_tree, layout="rectangular")
p2 <- ggtree(portal_tree, layout="ape")
p3 <- ggtree(portal_tree, layout="roundrect")
p4 <- ggtree(portal_tree, layout="slanted")
p5 <- ggtree(portal_tree, layout="ellipse")
p6 <- ggtree(portal_tree, layout="dendrogram") # pay attention to spelling on the option!
p7 <- ggtree(portal_tree, layout="circular")
p8 <- ggtree(portal_tree, layout="radial")
p9 <- ggtree(portal_tree, layout="fan", open.angle = 945)
p10 <- ggtree(portal_tree, layout="equal_angle")
p11 <- ggtree(portal_tree, layout="daylight")
p12 <- ggtree(portal_tree, layout="unrooted")
```

Part 2:
Create a subplot for each tree.
```{r}
plot_list(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12, tag_levels = "a")
```
Part 3: Is the tree representation rooted or unrooted?

```{r}
# create a vector of labels for my 12 plots:
my_labels <- paste(letters[1:12], "=", "rooted")
my_labels
plot_list(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12,
          labels = my_labels, tag_size = 5)
```


```{r}
taxonomy <- read.csv(file = "../data-raw/portal-species-taxonomy.csv")
species <- read.csv(file = "../data-raw/species.csv")
surveys <- read.csv(file = "../data-raw/surveys.csv")
```

Join the data table to a tree:

```{r}
intersect(colnames(species), colnames(surveys))
library(dplyr)
inner_join(species, surveys, by = "species_id")
```
```{r}
library(ape)
portal_tree <- read.tree(file = "../data-raw/portal-tree.tre")
portal_tree
class(portal_tree)
typeof(portal_tree)
str(portal_tree)
portal_tree
str(surveys)

```
```{r}
portal_tree <- read.tree(file = "../data-raw/portal-tree.tre")
portal_tree <- read.tree(file = "https://raw.githubusercontent.com/LunaSare/data-science-research-biology/main/data/portal-tree.tre")
portal_tree
# write phylo object to a newick file
write.tree(phy = portal_tree, file = "../data-raw/portal-tree-url.tre")

nrow(taxonomy)
taxonomy <- read.csv(file = "../data-raw/portal-species-taxonomy.csv")
taxonomy <- read.csv(file = "https://raw.githubusercontent.com/LunaSare/data-science-research-biology/main/data/portal-species-taxonomy.csv")
head(taxonomy)
write.csv(taxonomy, file = "../data-raw/taxonomy-test.csv", row.names = FALSE)
```

### Join the taxnomy data with the portal tree

```{r}
library(tidytree)
left_join(portal_tree, taxonomy, by = "label")
## the join failed!!
# we need to check our species labels on the tree and on the data table:
head(taxonomy)
taxonomy$label
```
Replace the tip labels from our original portal tree with a new vector of tip labels:
```{r}
tip_labels <- portal_tree$tip.label
tip_labels
tip_labels_cleaned <- gsub("_", " ", tip_labels)
portal_tree[["tip.label"]] <- tip_labels_cleaned
portal_tree$tip.label
```

```{r}
tree_data <- left_join(portal_tree, taxonomy, by = "label")
tree_data
class(tree_data)
```
Color by a variable in our column:
```{r}
library(ggtree)
ggtree(tr = tree_data, mapping = aes(color = taxa)) +
  geom_tiplab(size = 2.5, fontface = "italic") +
  xlim(0,100)
ggsave(filename = "../fig/portal-tree-1.png")
```

### Troubleshooting tree and data joins:

```{r}
portal_tree$tip.label
taxonomy$label
match(taxonomy$label, portal_tree$tip.label)
match(portal_tree$tip.label, taxonomy$label)
```

### Exercise - homework

Exercise: A taxonomy table for small_tree
1. Plot small tree with tip labels in italics and using the appropriate xlim
2. Find the appropriate scientific group labels for each genus in small_tree using this tree as guide.
Create a data frame with 3 columns:
a "label" column with the names of the tip labels of small_tree. Tip: extract the element "tip.label" from your phylo object to get a vector of tip labels that you can then join to the other vectors to create a data frame.
a "taxa" column with the scientific names of the group that each genus belongs to.
a "common_name" column with the common names of the group that each genus belongs to. Tip: use the function c() to create the vectors that will be columns "taxa" and "common_name"
Join your tree and your table using left_join().
Create two different tree plots using taxa and common name to color the tips of the tree.

```{r}
small_tree <- read.tree(file = "http://ape-package.ird.fr/APER/APER2/primfive.tre")
branching.times(small_tree)
ggtree(small_tree) + 
  xlim(0, 1.5) +
  geom_tiplab(fontface = "italic")

label <- small_tree$tip.label[1:2]
taxa <- c("Hominini", "Hominidae")
common_name <- c("Human", "Orangutan")

data.frame(label, taxa, common_name)


```
## Mapping data to the tree stucture

Load data from package TDbook using the function `data()`:
```{r}
install.packages("TDbook")
library(TDbook)
data("tree_boots")
data("df_tip_data")
data("df_inode_data")
tree_boots
# column "Newick_label" change to "label" in df_tip_data and in df_inode_data
```

Using mutate
```{r}
renamed_tip_data <- df_tip_data %>% 
  mutate(label = Newick_label)
colnames(renamed_tip_data)
renamed_inode_data <- df_inode_data %>% 
  mutate(label = newick_label)
colnames(renamed_inode_data)
renamed_inode_data$label
```
Using colnames function

```{r}
colnames(df_tip_data)[1] <- "label"
colnames(df_tip_data)
df_inode_data
colnames(df_inode_data)[1] <- "label"
colnames(df_inode_data)

```

Join `tree_boots` to `df_tip_data`:

```{r}
names(tree_boots)
colnames(df_tip_data)
intersect(names(tree_boots), colnames(df_tip_data))
tree_joined <- left_join(tree_boots, df_tip_data, by = "label")
tree_joined
```

Plot some data to the tips of `tree_boots`

```{r}
colnames(df_tip_data)
df_tip_data$rank

ggtree(tree_joined) +
  geom_tippoint(mapping = aes(shape = trophic_habit,
                              color = trophic_habit,
                              size = mass_in_kg)) +
  geom_tiplab(fontface = "italic") +
  xlim(0, 4)

df_tip_data

```

Exercise: Mapping weight data from `surveys` CSV table to the portal tree
1. Get the average weight and hindfoot length per species.
2. Create a new data frame that contains the taxonomy data plus the averaged data per species that you got on last question.
3. Create two plots with data on the tips, one with the average weight and the other with average hindfoot length. Make sure to also add tip labels, formatted in italics.

```{r}
library(dplyr)
colnames(surveys)
surveys %>% 
  filter(!is.na(weight)) %>% 
  group_by(species_id) %>% 
  summarise(mean_weight = mean(weight)) -> species_weight
nrow(species_weight)
spp_weight <- inner_join(species, species_weight)
spp_weight

colnames(taxonomy)
taxonomy_weight <- full_join(taxonomy, spp_weight)
taxonomy_weight

tree_join <- left_join(portal_tree, taxonomy_weight, by = "label")

tree_join

ggtree(tree_join) + 
  geom_tiplab(fontface = "italic", size = 2) +
  xlim(0, 20) +
  geom_tippoint(mapping = aes(shape = "circle",
                              size = mean_weight,
                              color = taxa))
```






