---
title: "small mammals dataset"
author: "Luna L Sanchez Reyes"
date: "2022-10-11"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The Portal Project Teaching Dataset

It is an ecological experiment made in the Chihuahuan desert in Arizona near the city of Portal.

Later I will add a picture of the Chihuahuan desert https://www.google.com/url?sa=i&url=https%3A%2F%2Fwww.desertusa.com%2Fchihuahuan-desert.html&psig=AOvVaw1ccqb81iiGHfvmvPe9xZXL&ust=1665611083984000&source=images&cd=vfe&ved=0CAwQjRxqFwoTCPj2uP2S2foCFQAAAAAdAAAAABAD

For all the data manipulations we are going to be using the `dplyr` package.

To install packages we use the function `install_packages()`. But to use the functions inside a package, we need to load them to our environment, we do this with the function `library()`.

```{r}
library(dplyr)
```

# Read the data table

```{r}
getwd()
surveys <- read.csv(file = "../data-raw/surveys.csv")
head(surveys)
colnames(surveys)
# View(surveys)
plots <- read.csv(file = "../data-raw/plots.csv")
head(plots)
species <- read.csv(file = "../data-raw/species.csv")
head(species)
```

# Subset a data table by columns

Using the dplyr function `select()`:

```{r}
select(surveys, year, species_id, weight)
surveys_short <- select(surveys, year, species_id, weight)
head(surveys_short)
```

Other ways to subset a data table by columns:

```{r}
columns2keep <- c("year", "species_id", "weight")
surveys[,columns2keep]
```

## Add columns to a data table


Let's say that we want to add an extra column that has weight divided by 10.

Using the classic approach

```{r}
colnames(surveys)
surveys$weight_decimal <- surveys$weight/10

colnames(surveys)
```
We can use the function `mutate()`

```{r}
surveys <- mutate(surveys, weight_decimal_2 = weight/10)
```

```{r}
surveys
```

## Rearrange/reorder a data table by some value

We will use the function `arange()`

```{r}
surveys_arranged <- arrange(surveys, weight)

tail(surveys_arranged, 100)
```
To arrange in a descending order:

```{r}
arrange(surveys, desc(weight))
```

## Filter data

To filter unwanted data or by some specific value in our data we use the function `filter()`

```{r}
filter(surveys, species_id == "DS")
filter(surveys, species_id != "DS")
```

We can also filter by multiple conditions:

```{r}
filter(surveys, species_id == "DS", year >= 1998)
# tail(temp_table, 100)
```

The logical operator `&` and `|` mean "AND" and "OR"

```{r}

filter(surveys, species_id == "DS" & year >= 1998)

```

```{r}
filter(surveys, species_id == "DS" | year >= 1998)
```

Let's filter for values that are NA, we nbeed the function `is.na()`

```{r}

is.na(surveys$weight)
surveys$weight

filter(surveys, !is.na(weight))
```


Exercise 1:

```{r}

# Question2:
subset1 <- select(surveys, year, month, day, species_id)
nrow(subset1)

# Question 3:
surveys_subset <- filter(surveys, !is.na(weight))
nrow(surveys_subset)
surveys_subset <- mutate(surveys_subset, weight_kg = weight/1000)
nrow(surveys_subset)
surveys_subset_filtered <- filter(surveys_subset, species_id == "SH")
nrow(surveys_subset_filtered)

surveys_subset <- select(surveys_subset, year, species_id, weight_kg)
```


```{r}
select(surveys, month, year, day)

surveys$weight == NA
NA == NA
# the following does not work:
#filter(surveys, match(surveys$weight, NA))
```

## Pipes

`%>%` - `magrittr` package required
`|>` - inbuilt in R, but very new, so you need an updated version of RStudio

```{r}
mean(x = c(1,3,65, 38, 100)) ##  sequential code
c(1, 3, 65, 38, 100) %>% mean()

sqrt(mean(x = c(1,3,65,38, 100)))
?sqrt
sqrt(41.4)
c(1,3,65,38, 100) |> mean () |> sqrt()

is.na(3)
!is.na(NA)
is.na(surveys$weight)

mean1 <- mean(x = c(1, 3, 7, 38, 100))
mean1

c(1, 3, 7, 38, 100) |> mean() -> mean2
mean2

mean3 <- c(1, 3, 7, 38, 100) |> mean()
mean3

mean(surveys$weight, na.rm = TRUE)
surveys$weight |> mean(na.rm = TRUE)
```

Exercise 2
Use data pipes to run the code from Exercise 1
```{r}
surveys1 <- surveys |> 
  mutate(weight_kg = weight/1000) |>
  select(year, species_id, weight_kg) |>
  filter(!is.na(weight_kg))
colnames(surveys1)
is.na(surveys1$weight_kg) |> any()

```

```{r}
surveys2 <- surveys |> 
  select(year, month, day, species_id, weight) |>
  filter(species_id == "SH") |>
  filter(!is.na(weight)) |>
  select(year, month, day, species_id)

colnames(surveys2)
nrow(surveys2)

surveys2$species_id == "SH"
surveys[surveys2$species_id == "SH",]
colnames(surveys2)
```

```{r}
ds_data <- filter(surveys, species_id == "DS", !is.na(weight))
nrow(ds_data)
ds_data_by_year <- arrange(ds_data, year)
ds_weight_by_year <- select(ds_data_by_year, year, weight, species_id)
nrow(ds_weight_by_year)
ds_weight_by_year
```

```{r, eval = FALSE}
ds_weight_by_year <- surveys |> 
  filter(species_id == "DS") |>
  filter(!is.na(weight)) |>
  arrange(year) |>
  select(year, weight, species_id)

ds_weight_by_year
xx <- lm(weight ~ year, data = surveys_DS)
summary(xx)
summary(lm(weight ~ year, data = surveys))
```

## Data aggregation

```{r}
surveys_by_year <- group_by(surveys, plot_id, year)   

surveys_by_year

plot_year_counts <- summarize(surveys_by_year, abundance = n())
```

The same but using a pipe:

```{r}
surveys_by_year <- surveys |>
  group_by(plot_id, year) |>
  summarize(abundance = n())

surveys_by_year
```
```{r}
surveys |> 
  filter(!is.na(weight)) |>
  filter(species_id == "DO") |>
  group_by(year) |>
  summarize(mean_by_year = mean(weight)) 
colnames(surveys)
```

```{r}
getwd()
read.csv(file = "../data-raw/shrub-volume-data.csv") -> shrubs
head(shrubs)

select(shrubs, length) -> shrubs_length
```

```{r}
filter(shrubs, height > 5)

1 == 1
1 > 1
1 >= 1
"a" == "a"
"a" == "A"
"a" != "A"
```





Logical statements, AND indicated by the & symbol, OR indicated the vertical bar "|" symbol 
```{r}
filter(shrubs, height > 5 | experiment == 2)
filter(shrubs, height < 5)
filter(shrubs, experiment ==1 | experiment == 3)
filter(shrubs, experiment >= 2)
```

Relational statements with missing values

```{r}
NA == 3
shrubs$height == NA
is.na(NA)
```

Function that summarize values ina vector like mean, max, min, sum
```{r}
mean(shrubs$height)
max(shrubs$height)
min(shrubs$height)
sum(shrubs$height, na.rm = TRUE)
sum(surveys$weight, na.rm = TRUE)
```

## Joining data tables

First, load the data, then look at it:

```{r}
species <- read.csv(file = "../data-raw/species.csv")
plots
```

```{r}
inner_join(surveys, species, by = "species_id")
left_join(surveys, species, by = "species_id")
right_join(surveys, species, by = "species_id")
full_join(surveys, species, by = "species_id")
```


How do we find out if we have matching columns in two tables

```{r}
surveys_colnames <- colnames(surveys)
species_colnames <- colnames(species)
surveys_colnames[7]
to_keep <- c(1, 4, 3, 2)
to_keep <- c(FALSE, TRUE, TRUE, FALSE)
species_colnames[to_keep]
```

The %in% operator

```{r}
1 %in% c(1, 2, 3)
c(1,2,3) %in% 1

matching_colnames <- colnames(species) %in% colnames(surveys)
matching_colnames
colnames(species)[matching_colnames]
```

How to join more than 2 tables?

```{r}
combined <- inner_join(surveys, species, by = "species_id") 
combined_full <- inner_join(combined, plots, by = "plot_id")
head(combined)
head(combined_full)
```
Run the code above using a pipe:
```{r}
combined_full <- surveys %>%
  inner_join(species, by = "species_id") %>%
  inner_join(plots, by = "plot_id")
head(combined_full)
```
Joint-exercise:

```{r}
surveys %>% 
  inner_join(plots, by = "plot_id") %>% 
  filter(plot_type == "Control") # gets 15,660 rows

filter(plots, plot_type == "Control") %>%
  inner_join(surveys, ., by = "plot_id") # gets 15,660 rows
  
```
Exercise 2:

```{r}
colnames(surveys)
combined_full <- inner_join(surveys, species, by = "species_id") %>% 
  inner_join(plots, by = "plot_id") 
colnames(combined_full)
combined_full %>% 
  filter(plot_type == "Control" | plot_type == "Long-term Krat Exclosure") %>% 
  filter(taxa == "Rodent") %>% 
    select(year, genus, species, weight, plot_type) %>% # 20,497 rows
  filter(!is.na(weight)) # 19,344 rows rows

  
#colnames(test_object)
#tail(test_object, 100)
#"Long-term Krat Exclosure" %in% test_object$plot_type
```


Exercise 3

```{r}
combined_full %>% 
  filter(species_id == "DO") %>% 
  select(year, month, day, species_id, weight) -> data1

combined_full %>% 
  filter(species_id == "PP" | species_id == "PB") %>% 
  filter(year > 1994) %>% 
  filter(!is.na(hindfoot_length)) %>% 
  select(year, species_id, hindfoot_length) -> data2

combined_full %>% 
  filter(!is.na(hindfoot_length)) %>% 
  group_by(species_id, year) %>% 
  summarize(mean_hindfoot_length = mean(hindfoot_length, na.rm = TRUE))

#m mean(c(NA, 3, 4,5), na.rm = TRUE)

surveys %>% 
  filter(!is.na(hindfoot_length)) %>% 
  group_by(species_id, year) %>% 
  summarize(avg = mean(hindfoot_length, na.rm = TRUE)) -> data3

combined_full %>% 
  filter(genus == "Dipodomys") %>% 
  select(year, genus, species, weight, plot_type)
```
```{r}
abc <- letters[c(1,2,3)]
abc[c(TRUE, FALSE)]
abc[FALSE]
seq(1, 100, 2)
??seq
c(1,2,3) %in% c(4,5,1) %>% 
  c(1,2,3)[.]
```

How to use the function `filter()`
Some examples of relational statements: they evaluate  relationships between two variables or objects
The symbols that we use are
```
== "equals to"
> "larger than"
< "smaller than"
!= "not equal to"
>= "larger or equal to"
<= "smaller or equal to"
```
```{r}
1 == 1
1 > 1
1 >= 1
1 != 1
1 < 2
1 <= 2
"luna" == "luna"
"luna" == "Luna"
"luna" < "Luna"
```

Logical statements test if two or more relational statements are true or false in combination.
To evaluate a logical comparison we have two symbols, that represent two types of logical comparisons:
- the ampersand &: this evaluates that the relation to the left AND the relation to the right are both TRUE
- the vertical bar |: evaluates that the relation to the left OR the relation to the right are TRUE; in other words, at least one of the relations must be TRUE

```{r}
1 == 1 & 1 < 2
1 == 1 | 1 < 2
1 == 1 | "luna" == "mars"
1 == 1 & "luna" == "mars"
```

```{r, eval = FALSE}
colnames(combined_full)
filter(combined_full, plot_type == "Control" & species_id == "PP" & plot_id == 2 | species_id == "DO") %>%  
  filter(data = ., !is.na(weight)) %>% 
  nrow(.)
```

Something important for relational statements is the NA
```{r}
NA == "luna"

is.na(NA)
is.na("luna")
is.na(3)

mean(surveys$weight, na.rm = TRUE)
min(surveys$weight, na.rm = TRUE)
max(surveys$weight, na.rm = TRUE)
```


```{r}
surveys$weight %>% 
  mean(na.rm = TRUE)
```

```{r}
surveys %>% 
  filter(!is.na(weight)) %>% 
  lm(weight ~ year, data = .)

surveys %>% 
  mutate(species_id = tolower(species_id))

intersect(colnames(surveys), colnames(plots))
intersect(colnames(species), colnames(surveys))
length <- c(2.2, 2.1, 2.7, 3.0, 3.1, 2.5, 1.9, 1.1, 3.5, 2.9)
width <- c(1.3, 2.2, 1.5, 4.5, 3.1, NA, 1.8, 0.5, 2.0, 2.7)
height <- c(9.6, 7.6, 2.2, 1.5, 4.0, 3.0, 4.5, 2.3, 7.5, 3.2)
length * width
```


```{r}
mutate(surveys, sex = tolower(sex))
```


```{r, child = 'vectors_and_data_frames.Rmd'}

````

```{r}

````

```{r}

````

```{r}

````