---
title: "small mammals dataset"
author: "Luna L Sanchez Reyes"
date: "2022-10-11"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The Portal Project Teaching Dataset

It is an ecological experiment made in the Chihuahuan desert in Arizona near the city of Portal.

Later I will add a picture of the Chihuahuan desert https://www.google.com/url?sa=i&url=https%3A%2F%2Fwww.desertusa.com%2Fchihuahuan-desert.html&psig=AOvVaw1ccqb81iiGHfvmvPe9xZXL&ust=1665611083984000&source=images&cd=vfe&ved=0CAwQjRxqFwoTCPj2uP2S2foCFQAAAAAdAAAAABAD

For all the data manipulations we are going to be using the `dplyr` package.

To install packages we use the function `install_packages()`. But to use the functions inside a package, we need to load them to our environment, we do this with the function `library()`.

```{r}
library(dplyr)
```

# Read the data table

```{r}
getwd()
surveys <- read.csv(file = "../data-raw/surveys.csv")
head(surveys)
colnames(surveys)
# View(surveys)
plots <- read.csv(file = "../data-raw/plots.csv")
head(plots)
species <- read.csv(file = "../data-raw/species.csv")
head(species)
```

# Subset a data table by columns

Using the dplyr function `select()`:

```{r}
select(surveys, year, species_id, weight)
surveys_short <- select(surveys, year, species_id, weight)
head(surveys_short)
```

Other ways to subset a data table by columns:

```{r}
columns2keep <- c("year", "species_id", "weight")
surveys[,columns2keep]
```

## Add columns to a data table


Let's say that we want to add an extra column that has weight divided by 10.

Using the classic approach

```{r}
colnames(surveys)
surveys$weight_decimal <- surveys$weight/10

colnames(surveys)
```
We can use the function `mutate()`

```{r}
surveys <- mutate(surveys, weight_decimal_2 = weight/10)
```

```{r}
surveys
```

## Rearrange/reorder a data table by some value

We will use the function `arange()`

```{r}
surveys_arranged <- arrange(surveys, weight)

tail(surveys_arranged, 100)
```
To arrange in a descending order:

```{r}
arrange(surveys, desc(weight))
```

## Filter data

To filter unwanted data or by some specific value in our data we use the function `filter()`

```{r}
filter(surveys, species_id == "DS")
filter(surveys, species_id != "DS")
```

We can also filter by multiple conditions:

```{r}
filter(surveys, species_id == "DS", year >= 1998)
# tail(temp_table, 100)
```

The logical operator `&` and `|` mean "AND" and "OR"

```{r}

filter(surveys, species_id == "DS" & year >= 1998)

```

```{r}
filter(surveys, species_id == "DS" | year >= 1998)
```

Let's filter for values that are NA, we nbeed the function `is.na()`

```{r}

is.na(surveys$weight)
surveys$weight

filter(surveys, !is.na(weight))
```


Exercise:

```{r}

# Question2:
subset1 <- select(surveys, year, month, day, species_id)
nrow(subset1)

# Question 3:
surveys_subset <- filter(surveys, !is.na(weight))
nrow(surveys_subset)
surveys_subset <- mutate(surveys_subset, weight_kg = weight/1000)
nrow(surveys_subset)
surveys_subset_filtered <- filter(surveys_subset, species_id == "SH")
nrow(surveys_subset_filtered)

surveys_subset <- select(surveys_subset, year, species_id, weight_kg)
```


```{r}
select(surveys, month, year, day)

surveys$weight == NA
```

