---
title: "uhuru dataset"
author: "Luna L Sanchez Reyes"
date: "2022-10-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      root.dir = "/Users/lunasare/Desktop/data-science-course/fall-2022/")

```

### 1. Describing the data that we are using

We are using the dataset from [this study](https://esapubs.org/archive/ecol/E095/064/metadata.php)

Add a picture of an Acacia

### 2. Reading the data table into R

First make sure that we are in the correct working directory, we use function `getwd()`

```{r}
getwd()
```
If it is not correct set the working directory in the setup chunk above


```{r}
acacia <- read.csv(file = "/Users/lunasare/Desktop/data-science-course/fall-2022/data-raw/ACACIA_DREPANOLOBIUM_SURVEY.txt", sep = "\t")
```
### 3. Explore our data set

```{r}
head(acacia)
summary(acacia)
colnames(acacia)

```

Make sure that everuthing that is a number, is actually numeric.

One way to do this is withthe function `summary`, and checking at the type of data on each column visually.

Another way is using the type function

```{r}
typeof(acacia[,"HEIGHT"])

acacia$HEIGHT
```

We identified a column that has problematic data. We need to fix it!

We are goinf to read the data table again, but we are gonna assign `NA` to the "dead" value that we do not want in our "HEIGHT" column.

```{r}
acacia <- read.csv(file = "/Users/lunasare/Desktop/data-science-course/fall-2022/data-raw/ACACIA_DREPANOLOBIUM_SURVEY.txt", 
                   sep = "\t", 
                   na.strings = "dead")

```


Let's check if this worked!

```{r}
acacia$HEIGHT
typeof(acacia$HEIGHT)
```


### 4. Visualize our data

For this, we are using the `ggplot` package.
Let's install it and load it:

```{r}
# install.packages("ggplot2")
library(ggplot2)
```

Now we are gonna create our first plotting layer with the function `ggplot`. 

```{r}
colnames(acacia)
acacia$CIRC
```

```{r}
ggplot(data = acacia, mapping = aes(x = CIRC, y = HEIGHT)) +
  geom_point(size = 1, color = "blue", alpha = 0.1) +
  labs(x = "Circumference", y = "Height", title = "Data from UHURU Acacia survey")


```
To rescale the plotting of the axis to log scale we use the function `scale_y_log_10()`

```{r}
ggplot( data = acacia, mapping = aes(x = CIRC, y = HEIGHT)) +
  geom_point(size =1, color = "purple", alpha = 0.4) +
  scale_x_log10() +
  scale_y_log10()
```
We have the information on experimental treatment in th etreatment column:

```{r}
acacia$TREATMENT
```

Let's add information on treatment to our plot:

```{r}
ggplot(data = acacia, mapping = aes (x = CIRC, y = HEIGHT, color = TREATMENT)) +
  geom_point(size = 2, alpha = 0.5)

```


### 4.2 Visualize a statistical analysis of correlation


