---
title: "uhuru dataset"
author: "Luna L Sanchez Reyes"
date: "2022-10-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Describing the data that we are using

We are using the dataset from [this study](https://esapubs.org/archive/ecol/E095/064/metadata.php)

Add a picture of an Acacia

### 2. Read the data set into R

#### 2.1 Get the working directory of my R chunks

The working directory of an Rmd file is not the same as the working directory of the Rstudio project you are in. We can check that by running the following.

How do I test where the computer is?

```{r}
r_proj_wd <- "/Users/lunasare/Desktop/data-science-course/fall-2022"
r_chunk_wd <- getwd()
r_proj_wd == r_chunk_wd
r_chunk_wd
```

As you can see, the working directory from my R project and the one used in my R chunks in an Rmd file are not the same.

The working directory of an R chunk is equal to the path where the Rmd file is in, in my case, it is `r_chunk_wd`. This is nested withing my project directory, keep this in mind for when you have to provide the path of a file ;)

#### 2.2 Read the data set into R and create an object

```{r}
# getwd()
acacia <- read.csv(file = "../data-raw/ACACIA_DREPANOLOBIUM_SURVEY.txt", sep = "\t")
acacia
```

### 3. Explore the data set

```{r}
head(acacia)
summary(acacia)
colnames(acacia)
rownames(acacia)
class(acacia$SURVEY)
# the sapply function allows to apply a function to a list of objects
# a data frame is a list of vectors of the same length
sapply(acacia, class)
View(acacia)
```

Make sure that everything that is a number is actually `typeof` numeric or double or integer.

One way to do this is with the function `summary`, and checking at the type of data on each column visually.

Another way is using the `typeof()` function:

```{r}
typeof(acacia[,"HEIGHT"])

acacia$HEIGHT
```

We identified a column that has problematic data. We need to fix it!

## Cleaning our raw data

### 3.2 Assigning `NA` to missing/incorrect values

We are going to read the data table again, but we are gonna assign `NA` to the "dead" value that we do not want in our "HEIGHT" column.

```{r}
acacia <- read.csv(file = "../data-raw/ACACIA_DREPANOLOBIUM_SURVEY.txt", 
                   sep = "\t", 
                   na.strings = "dead")

```

Let's check if this worked:

```{r}
acacia$HEIGHT
typeof(acacia$HEIGHT)
class(acacia$HEIGHT)
```

### 4. Visualize our data

For this, we are using the `ggplot2` package. Let's install it and load it:

```{r}
# install.packages("ggplot2")
library(ggplot2)
```

Now we are gonna create our first plotting layer with the function `ggplot`.

```{r}
colnames(acacia)
acacia$CIRC
```

The ggplot function creates a blank canvas. The canvas contains our data. and the variables that we want to plot.

A scatter plot can be created with the function `geom_point()`

```{r}
colors()
```

```{r}
?pch
```

```{r}
ggplot(data = acacia, mapping = aes(x = CIRC, y = HEIGHT)) +
  geom_point(size = 1, color = "tomato", alpha = 0.8, shape = 11) +
  labs(x = "Circumference", y = "Height", title = "Data from UHURU Acacia survey")
```

To rescale the plotting of the axis to log scale we use the function `scale_y_log_10()` and `scale_x_log10()`:

```{r}
ggplot( data = acacia, mapping = aes(x = CIRC, y = HEIGHT)) +
  geom_point(size =1, color = "purple", alpha = 0.4) +
  scale_x_log10() +
  scale_y_log10()
```

## Grouping data:

We have the information on experimental treatment in the column named "TREATMENT":

```{r}
acacia$TREATMENT
```

Let's add information about treatment to our plot:

```{r}
ggplot(data = acacia, mapping = aes (x = CIRC, y = HEIGHT, color = TREATMENT)) +
  geom_point(size = 2, alpha = 0.5)

```

# Subplots or facets

The function to create subplots by a third variable is called `facet_wrap()`

```{r}
ggplot(data = acacia, mapping = aes(x = CIRC, y = HEIGHT, color = TREATMENT)) +
  geom_point() +
  facet_wrap(~TREATMENT)
```

A little interpretation of what is going on here...

How to test these hypotheses?

## Model fitting functions

The `geom_smooth()` function alloes to fit linear models to a set of points

```{r}
ggplot(data = acacia, mapping = aes(x = CIRC, y = HEIGHT, color = TREATMENT, linetype = TREATMENT)) +
  geom_point() +
  geom_smooth(method = "loess") + 
  facet_wrap(~TREATMENT)
```

## Histograms and barplots

For bar plots we use the geom_bar() function:

```{r}
ggplot(data = acacia, aes(x = TREATMENT)) +
  geom_bar()
```

If we want to see the distribution of a continuous variable we use the geom_histogram() function:

```{r}
ggplot(data = acacia, mapping = aes(x = HEIGHT, fill = TREATMENT)) +
  geom_histogram(bins = 20, color = "black")
```

## Layer multiple data from the same or different data sets

```{r}
ggplot() +
  geom_point(data = acacia,
             mapping = aes(x = CIRC, y = HEIGHT,
                           color = TREATMENT)) +
  geom_smooth(data = acacia, 
              mapping = aes(x = CIRC, y = HEIGHT)) +
  geom_histogram(data = acacia, 
                 mapping = aes(x = CIRC, color = TREATMENT), alpha = 0.2)

```

# Save images as files

```{r}
ggsave(filename = "acacia_by_treatment.pdf", width = 5, height = 3)
```

# Exercise 1

```{r}
ggplot(data = acacia, mapping = aes(x = CIRC, y = AXIS1, color = ANT)) +
  geom_point() +
  labs(x = "Circumference", y = "Canopy Diameter") +
  facet_wrap(~ANT) +
    geom_smooth()
```

Exercise 2.

```{r}
ggplot(data = acacia, mapping = aes(x = ANT)) +
  geom_bar()
```
```{r}
ggplot(data = acacia, mapping = aes(x = HEIGHT)) +
  geom_histogram(binwidth = 0.08) +
  labs(x = "Height (m)", y = "Number of Acacia")
```


```{r}
ggplot(data = acacia) +
  geom_histogram(mapping = aes(x = AXIS1), 
                 fill = "red", alpha = 0.3, bins = 10) +
  geom_histogram(mapping = aes(x = AXIS2), 
                 fill = "black", alpha = 0.3, bins = 10) +
  facet_wrap(~TREATMENT)
```


Exercise 3:

```{r}
library(readr)
trees <- read_tsv("../data-raw/TREE_SURVEYS.txt",
               col_types = list(HEIGHT = col_double(),
                                AXIS_2 = col_double()))
View(trees)
```

```{r}
trees$canopy_area <- trees$AXIS_1 * trees$AXIS_2

trees_subset <- trees[, c("SURVEY", "YEAR", "SITE", "canopy_area")]
head(trees_subset)
```

```{r}
colnames(trees)
trees$canopy_area

ggplot(data = trees, mapping = aes(x= canopy_area, y = HEIGHT), color = TREATMENT) +
  geom_point() + 
  facet_wrap(~SPECIES)
```

```{r}
which(trees$AXIS_1 > 20)
which(trees$AXIS_2 > 20)
trees[3839]
trees[(trees$AXIS_2 > 20),]

which(trees$AXIS_1 > 20)
which(trees$AXIS_2 > 20)
to_remove <- which(trees$AXIS_1 > 20 | trees$AXIS_2 > 20)

trees <- trees[-to_remove,]
to_remove_again <- which(is.na(trees$SPECIES))
trees <- trees[-to_remove_again,]
```

```{r}
ggplot(data = trees, mapping = aes(x= canopy_area, y = HEIGHT, color = TREATMENT)) +
  geom_point() +
  facet_wrap(~SPECIES)
```

```{r}
library(dplyr)
trees %>%
  group_by(YEAR, SPECIES) %>%
  summarize(species_abundance = n()) -> abundance

abundance
```
```{r}
ggplot(data = abundance, mapping = aes(x = YEAR, y = species_abundance)) +
  geom_point() + 
  geom_line() +
  facet_wrap(~SPECIES, scales = "free_y")

test1 <- trees$SPECIES == "Croton" & trees$YEAR == 2013
sum(test1)
trees[which(test1),]
```


Exercise 4

```{r}
ggplot() + 
  geom_point(data = trees, mapping = aes(x= CIRC, y = HEIGHT), color = "gray", alpha = 0.3) +
  geom_smooth(data = trees, mapping = aes(x= CIRC, y = HEIGHT), color = "black", method = "lm") +
  scale_x_log10() +
  scale_y_log10() +
  geom_point(data = acacia, mapping = aes(x= CIRC, y = HEIGHT), color = "red", alpha = 0.3) +
  geom_smooth(data = acacia, mapping = aes(x= CIRC, y = HEIGHT), color = "red", method = "lm") +
  scale_x_log10() +
  scale_y_log10()
  
```